---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

We will unzip the activity file, and load the resulting csv as a data frame. 

```{r}
unzip("activity.zip")
df <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?

Here's a histogram of the total number of steps taken each day. 
```{r}
library(dplyr, warn.conflicts = FALSE)
df1 <- df %>% group_by(date) %>% summarize(total_steps = sum(steps, na.rm = TRUE))
hist(df1[, 'total_steps'])
```

The table below shows the mean/median total number of steps taken per day. On days `where no measurements were available, you will see the mean/median as NaN/NA. 

```{r}
df %>% group_by(date) %>% summarize(avg_steps = mean(steps, na.rm = TRUE), median_steps = median(steps, na.rm = TRUE))
```


## What is the average daily activity pattern?

The following shows a time-series plot, with each 5-minute interval on the x-axis, and the average number of steps in that interval across all days on the y-axis. 

```{r}
df1 <- df %>% group_by(interval) %>% summarize(avg_steps = mean(steps, na.rm = TRUE))
plot(df1$interval, df1$avg_steps, type = 'l')
interval <- df1[df1$avg_steps == max(df1$avg_steps), ]$interval
```

The 5-minute interval with the highest average is `r interval`. 

## Imputing missing values

The number of missing values for the steps parameter is `r sum(is.na(df$steps))`. 
```{r}
sum(is.na(df$steps))
```

We will fill in each missing value with the corresponding mean for the 5-minute interval. With this strategy, here's the new data set. 
```{r}
df2 <- df
df1 <- df %>% group_by(interval) %>% summarize(avg_steps = mean(steps, na.rm = TRUE))
indices <- which(is.na(df2$steps))
for(i in indices) {
  df2[i, 'steps'] <- df1[df1$interval == df2[i, 'interval'], 'avg_steps']
}
```

Here's the mean and median with the new imputed missing values. Only the days with missing values are now impacted; all the NaNs and NAs are eliminated. 

```{r}
df2 %>% group_by(date) %>% summarize(avg_steps = mean(steps, na.rm = TRUE), median_steps = median(steps, na.rm = TRUE))
```

## Are there differences in activity patterns between weekdays and weekends?

Let's add a new factor variable which identifies whether the day is a weekday or a weekend. 
```{r}
df2$weekday <- factor(ifelse(weekdays(as.Date(df2$date)) %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"), "Weekday", "Weekend"))
```

Here's a histogram of the total number of steps taken each day after the missing values are imputed. 
```{r}
df1 <- df2 %>% group_by(date) %>% summarize(total_steps = sum(steps, na.rm = TRUE))
hist(df1[, 'total_steps'])
```

Here's a panel plot showing time series with the 5-minute intervals on the x-axis, and the average number of steps on the y-axis. 

```{r}
df3 <- df2 %>% group_by(interval, weekday) %>% summarize(avg_steps = mean(steps, na.rm = TRUE))
df4 <- df3[df3$weekday == "Weekday", ]
df5 <- df3[df3$weekday == "Weekend", ]

library(lattice)
layout(c(1, 2), 2, 1)
plot(df4$interval, df4$avg_steps, type = "l", main = "Weekday")
plot(df5$interval, df5$avg_steps, type = "l", main = "Weekend")
```